---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# `artcatr`

`artcatr` calculates sample size and power for a two-arm randomized trial with an ordinal outcome, analysed by a proportional-odds model. It uses the `ologit` method of White, Marley-Zagar, Morris, Parmar, Royston & Babiker (2023). This package implements the Stata functions (`artcat`, White et al. (2023)) into R (`artcatr`).


## Installation

`artcatr` can be installed via 

```{r, eval = FALSE}
install.packages('artcatr', repos = c('https://dcr-unibe-ch.r-universe.dev', 'https://cloud.r-project.org'))
```

or via

```{r, eval = FALSE}
remotes::install_github("dcr-unibe-ch/artcatr")
```

This may require `Sys.setenv(R_REMOTES_NO_ERRORS_FROM_WARNINGS="true")` if packages were built under a different R version to the one you are using.

## Usage

```{r, message=FALSE}
library(artcatr)
```


